---
title: "Bayesian final study"
author: "GCSM1"
date: "2024-08-16"
output: html_document
---

## Load packages and libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(sf)
library(rstan)
library(bayesplot)
library(ggplot2)
library(spdep)
library(readr)
```

## Load data
```{r}
religion_2001 <- read_csv("C:/Users/USUARIO/Documents/UCL//Dissertation/v3 longitudinal diaspora study/Data/neighborhood data/religion borough 2001.csv")
religion_2011 <- read_csv("C:/Users/USUARIO/Documents/UCL//Dissertation/v3 longitudinal diaspora study/Data/neighborhood data/religion lsoa 2011.csv")
religion_2021 <- read_csv("C:/Users/USUARIO/Documents/UCL//Dissertation/v3 longitudinal diaspora study/Data/neighborhood data/religion msoa 2021.csv")
greenery_df <- read_csv("C:/Users/USUARIO/Documents/UCL//Dissertation/v3 longitudinal diaspora study/Data/neighborhood data/greenery boroughs.csv")
crime_df <- read_csv("C:/Users/USUARIO/Documents/UCL//Dissertation/v3 longitudinal diaspora study/Data/neighborhood data/historic crime 2010-2021.csv")
accomodation_2011 <- read_csv('C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/housing 2011.csv')
accomodation_2021 <- read_csv('C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/housing 2021.csv')
tenure_2001 <- read_csv('C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/home tenure 2001.csv')
tenure_2011 <- read_csv('C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/home tenure 2011.csv')
tenure_2021 <- read_csv('C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/home tenure 2021.csv')
housing_data <- read_csv('C:/Users/USUARIO/Documents/UCL/Summer/bayesian/housing_in_london_yearly_variables.csv')

# Spatial data
msoa <- st_read("C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/MSOA_2011_London_gen_MHW.shp")
lsoa <- st_read("C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/LSOA_2011_London_gen_MHW.shp")
borough <- st_read("C:/Users/USUARIO/Documents/UCL/Dissertation/v3 longitudinal diaspora study/Data/London_Borough_Excluding_MHW.shp")
```

## Clean religion data
```{r}
# Religion 2001
library(dplyr)

# Convert the columns representing religions to numeric, if necessary
religion_2001[ , 4:11] <- lapply(religion_2001[ , 4:11], as.numeric)

names(religion_2001)

#Eliminate columns
religion_2001 <- religion_2001[, -c(1, 3)]

# Define the function to find the majority religion
majority_religion <- function(row) {
  # Extract only the numeric values
  religion_counts <- c(
    Christian = row["Christian"],
    Buddhist = row["Buddhist"],
    Hindu = row["Hindu"],
    Jewish = row["Jewish"],
    Muslim = row["Muslim"],
    Sikh = row["Sikh"],
    `Other religions` = row["Other religions"],
    `No religion` = row["No religion"]
  )
  
  # Find the maximum value
  max_count <- max(religion_counts)
    # Find the names with the maximum value
  majority_religions <- names(religion_counts)[religion_counts == max_count]
  
  # Check for ties
  if (length(majority_religions) > 1) {
    return("No single majority")
  }
  
  return(majority_religions)
}


# Apply the function to each row
religion_2001$Majority_Religion <- apply(religion_2001, 1, majority_religion)

#Eliminate columns
religion_2001 <- religion_2001[, -c(2, 3, 4, 5, 6, 7, 8, 9)]

merged_2001_religion <- religion_2001 %>%
  left_join(msoa, by = c("Borough" = "LAD11NM"))

#Eliminate columns
merged_2001_religion <- merged_2001_religion[, -c(1, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)]
```

```{r}
# Religion 2011

#Eliminate columns
religion_2011 <- religion_2011[, -c(2)]

# Rename columns for easier access
religion_2011 <- religion_2011 %>%
  rename(
    LSOA = `Lower Super Output Area`,
    Christian = `Religion;Christian;2011`,
    Buddhist = `Religion;Buddhist;2011`,
    Hindu = `Religion;Hindu;2011`,
    Jewish = `Religion;Jewish;2011`,
    Muslim = `Religion;Muslim;2011`,
    Sikh = `Religion;Sikh;2011`,
    Other_religion = `Religion;Other religion;2011`,
    No_religion = `Religion;No religion;2011`,
    Religion_not_stated = `Religion;Religion not stated;2011`
  )

# Convert relevant columns to numeric
religion_2011 <- religion_2011 %>%
  mutate(across(Christian:Religion_not_stated, as.numeric))

# Revised function to determine the majority religion
majority_religion <- function(Christian, Buddhist, Hindu, Jewish, Muslim, Sikh, Other_religion, No_religion, Religion_not_stated) {
  # Create a named vector of religion counts
  religion_counts <- c(
    Christian = Christian,
    Buddhist = Buddhist,
    Hindu = Hindu,
    Jewish = Jewish,
    Muslim = Muslim,
    Sikh = Sikh,
    Other_religion = Other_religion,
    No_religion = No_religion,
    Religion_not_stated = Religion_not_stated
  )
  
  # Remove NA values
  religion_counts <- na.omit(religion_counts)
  
  # Check if there are any valid counts left
  if (length(religion_counts) == 0) {
    return("No data available")
  }
  
  # Find the maximum count
  max_count <- max(religion_counts)
  
  # Identify which religions have the maximum count
  majority_religions <- names(religion_counts)[religion_counts == max_count]
  
  # Return the majority religion if there’s a clear winner
  if (length(majority_religions) == 1) {
    return(majority_religions)
  } else {
    return("No single majority")
  }
}

library(dplyr)

# Apply the function to each row
religion_2011 <- religion_2011 %>%
  rowwise() %>%
  mutate(Majority_Religion = majority_religion(
    Christian, Buddhist, Hindu, Jewish, Muslim, Sikh, Other_religion, No_religion, Religion_not_stated
  )) %>%
  ungroup()

#Eliminate columns
religion_2011 <- religion_2011[, -c(2, 3, 4, 5, 6, 7, 8, 9, 10)]

# Define a function to get the most frequent religion
most_frequent_religion <- function(religions) {
  # Count the occurrences of each religion
  counts <- table(religions)
  
  # Find the religion with the highest count
  max_count <- max(counts)
  most_frequent_religions <- names(counts)[counts == max_count]
  
  # Return the most frequent religion if there’s a clear winner
  if (length(most_frequent_religions) == 1) {
    return(most_frequent_religions)
  } else {
    return("No single majority")
  }
}

merged_2011_religion <- religion_2011 %>%
  left_join(lsoa, by = c("LSOA" = "LSOA11CD"))

# Aggregate by MSOA and determine the most frequent majority religion
aggregated_2011_religion <- merged_2011_religion %>%
  group_by(MSOA11CD) %>%  # Use MSOA11CD or MSOA11NM for grouping
  summarise(
    Most_Frequent_Religion = most_frequent_religion(Majority_Religion),
    .groups = 'drop'  # Drop the grouping structure
  )

# Join 2011 and 2001 religion
religion_joined <- merged_2001_religion %>%
  left_join(aggregated_2011_religion, by = "MSOA11CD")

# Rename columns using rename()
religion_joined  <- religion_joined  %>%
  rename(
    `2001` = `Majority_Religion`,
    `2011` = `Most_Frequent_Religion`,
  )
```

```{r}
# Religion 2021

# Rename the column 'Middle layer Super Output Areas Code' to 'msoa11cd'
religion_2021 <- religion_2021 %>%
  rename(msoa11cd = `Middle layer Super Output Areas Code`)

# Print column names to verify
print(names(religion_2021))

# Filter out 'Does not apply' and find the majority religion for each MSOA
majority_religion <- religion_2021 %>%
  filter(`Religion (detailed) (58 categories)` != "Does not apply") %>%
  group_by(msoa11cd) %>%
  slice(which.max(Observation)) %>%
  dplyr::select(msoa11cd, `Religion (detailed) (58 categories)`) %>%
  rename(religion = `Religion (detailed) (58 categories)`)

# join to shp
final_religion <- religion_joined %>%
  left_join(majority_religion, by = c("MSOA11CD" = "msoa11cd")) 

# Rename columns 
final_religion  <- final_religion  %>%
  rename(
    `2021` = `religion`,
  )

# Create a data frame with years 2011-2018
years <- data.frame(year = 2011:2018)

# Merge files
expanded_religion <- merge(majority_religion, years)
```

## Clean greenery data
```{r}
library(dplyr)

greenery_df_cleaned <- greenery_df %>%
  rename(area = lb_code) %>%  # Rename lb_code to area
  dplyr::select(area, percent_green) %>%  # Select the desired columns
  na.omit() %>%  # Remove rows with NA values
  mutate(percent_green = as.numeric(percent_green))

# Create a data frame with years 
years <- seq(2011, 2018)

# Use merge() to perform a Cartesian join
expanded_grenery <- merge(greenery_df_cleaned, years)

# Change column name
names(expanded_grenery)[names(expanded_grenery) == "y"] <- "Year"

# Join to msoa chp
borough_greenery <- expanded_grenery %>%
  left_join(msoa, by = c("area" = "LAD11NM"))

# Calculate greenery of each msoa
borough_pop <- borough_greenery %>%
  group_by(LAD11CD) %>%
  summarize(total_borough_population = sum(USUALRES))

# Merge total borough population back to the original dataframe
borough_greenery <- merge(borough_greenery, borough_pop, by = "LAD11CD")

# Calculate the proportion of each MSOA in the borough
borough_greenery$population_proportion <- borough_greenery$USUALRES / borough_greenery$total_borough_population

# Assign greenery percentage to each MSOA
borough_greenery$MSOA_percent_green <- borough_greenery$population_proportion * borough_greenery$percent_green

# Select relevant columns for the result
msoa_green <- dplyr::select(borough_greenery, MSOA11CD, MSOA11NM, Year, MSOA_percent_green)
```
## Clean crime data
```{r}
library(dplyr)
library(tidyr)
# Crime data

# Rename 'Borough' column to 'Code'
crime_df <- crime_df %>% rename(code = Borough)

# Aggregate crime values by 'Code'
aggregated_crime <- crime_df %>%
  group_by(code) %>%
  summarise(across(starts_with("20"), sum, na.rm = TRUE))

# Sum values by year
aggregated_crime <- aggregated_crime %>%
  gather(key = "YearMonth", value = "Count", -code) %>%
  mutate(Year = substr(YearMonth, 1, 4)) %>%
  group_by(code, Year) %>%
  summarise(Yearly_Total = sum(Count)) %>%
  spread(key = Year, value = Yearly_Total)

# Filter to keep only years 2011-2018
years_to_keep <- as.character(seq(2011, 2018, by = 1))
final_crime <- dplyr::select(aggregated_crime, code, all_of(years_to_keep))

# Transform to long format
long_crime <- final_crime %>%
  pivot_longer(cols = `2011`:`2018`, names_to = "Year", values_to = "crime_counts")

# Join to msoa chp
borough_crime <- long_crime %>%
  left_join(msoa, by = c("code" = "LAD11CD"))

# Calculate crime of each msoa
borough_pop <- borough_crime %>%
  group_by(code) %>%
  summarize(total_borough_population = sum(USUALRES))

# Merge total borough population back to the original dataframe
borough_crime <- merge(borough_crime, borough_pop, by = "code")

# Calculate the proportion of each MSOA in the borough
borough_crime$population_proportion <- borough_crime$USUALRES / borough_crime$total_borough_population

# Assign percentage to each MSOA
borough_crime$MSOA_crime <- borough_crime$population_proportion * borough_crime$crime_counts

# Select relevant columns for the result
msoa_crime <- dplyr::select(borough_crime, MSOA11CD, MSOA11NM, Year, MSOA_crime)
```

## Clean accommodation data
```{r}
# Accommodation 2011

# Select columns 'Area' and 'Overcrowded ( -1 or less)' from 'house_2011'
accomodation_2011 <- accomodation_2011[, c('Area', 'Overcrowded ( -1 or less)')]

# Calculate total population per borough
msoa <- msoa %>%
  group_by(LAD11NM) %>%
  mutate(TotalPopulation = sum(USUALRES))

# Calculate the proportion for each MSOA
msoa <- msoa %>%
  mutate(Proportion = USUALRES / TotalPopulation)

# Merge accommodation2011 data with MSOA data
msoa_accomodation_2011 <- merge(msoa, accomodation_2011, by.x = "LAD11NM", by.y = "Area")

# Calculate distributed data
msoa_accomodation_2011 <- msoa_accomodation_2011 %>%
  mutate(DistributedOvercrowded = Proportion * `Overcrowded ( -1 or less)`)

# Select final columns
final_accomodation_2011 <- msoa_accomodation_2011 %>%
  dplyr::select(MSOA11NM, MSOA11CD, geometry, DistributedOvercrowded)

final_accomodation_2011 <- final_accomodation_2011 %>%
  rename(
    occupation_2011 = DistributedOvercrowded
  )

# Accommodation 2021

# Selecting columns for accommodation_2021 dataframe
accomodation_2021 <- accomodation_2021[, c('Lower layer Super Output Areas Code',
                             'Lower layer Super Output Areas',
                             'Occupancy rating for bedrooms (5 categories)', 'Observation')]

# Check the structure of the dataframe to ensure column names are correct
str(accomodation_2021)

# Filter the rows with occupancy rating of bedrooms: "-1 or less"
clean_accomodation_2021 <- accomodation_2021 %>%
  group_by(`Lower layer Super Output Areas Code`) %>%
  filter(row_number() %in% 33:40) %>%
  ungroup()

# Summarize the total values for each "Lower layer Super Output Areas Code"
aggregated_accomodation_2021 <- clean_accomodation_2021 %>%
  group_by(`Lower layer Super Output Areas Code`, `Lower layer Super Output Areas`, `Occupancy rating for bedrooms (5 categories)`) %>%
  summarise(Total_Observations = sum(Observation, na.rm = TRUE)) %>%
  ungroup()

# Merge to the LSOA codes
merged_accomodation_2021 <- merge(aggregated_accomodation_2021, lsoa, by.x = "Lower layer Super Output Areas Code", by.y = "LSOA11CD")

# Aggregate at MSOA level
agg_accomodation_2021 <- merged_accomodation_2021 %>%
  group_by(MSOA11CD, MSOA11NM) %>%
  summarise(
    Total_Observations = sum(Total_Observations),
    USUALRES = sum(USUALRES),
    HHOLDRES = sum(HHOLDRES),
    COMESTRES = sum(COMESTRES),
    POPDEN = mean(POPDEN),
    HHOLDS = sum(HHOLDS),
    AVHHOLDSZ = mean(AVHHOLDSZ)
  )

# Select final columns
final_accomodation_2021 <- agg_accomodation_2021 %>%
  dplyr::select(MSOA11NM, MSOA11CD, 'Total_Observations')

# Merge to the MSOA codes
ffinal_accomodation_2021 <- merge(final_accomodation_2021, msoa, by = "MSOA11NM")

# Select final columns
fffinal_accomodation_2021 <- ffinal_accomodation_2021 %>%
  dplyr::select(MSOA11NM, MSOA11CD.x, 'Total_Observations', geometry)

fffinal_accomodation_2021 <- fffinal_accomodation_2021 %>%
  rename(
    occupation_2021 = 'Total_Observations'
  )

# Total accommodation dataset

# Join both datasets
accommodation <- final_accomodation_2011 %>%
  inner_join(fffinal_accomodation_2021, by = c("MSOA11NM", "MSOA11CD" = "MSOA11CD.x", "geometry"))

# Interpolate missing years
# Convert the geometry column to an sf object
accommodation <- st_as_sf(accommodation, crs = 27700)

# Reshape the data for easier interpolation
accommodation_long <- accommodation %>%
  pivot_longer(
    cols = c(occupation_2011, occupation_2021), 
    names_to = "Year", 
    names_prefix = "occupation_", 
    values_to = "Occupation"
  ) %>%
  mutate(Year = as.numeric(Year))

# Define years to interpolate
years <- seq(2011, 2018)

# Perform interpolation for each MSOA
interpolated_accommodation <- accommodation_long %>%
  group_by(MSOA11NM, MSOA11CD, geometry) %>%
  do({
    data.frame(Year = years,
               Occupation = approx(.$Year, .$Occupation, xout = years, rule = 2)$y)
  }) %>%
  ungroup() %>%
  st_as_sf()
```

```{r}
# Preprocess Tenure data

# Calculate proportions 2001
tenure_2001 <- tenure_2001 %>%
  mutate(`2001 Owned` = `Tenure: Owned: Total; measures: Value` / `Tenure: All categories: Tenure; measures: Value`,
         `2001 Rented` = `Tenure: Private rented: Total; measures: Value` / `Tenure: All categories: Tenure; measures: Value`,
         `2001 Social` = `Tenure: Social rented: Total; measures: Value` / `Tenure: All categories: Tenure; measures: Value`)

# Keep only the required columns 2001
tenure_2001 <- tenure_2001 %>%
  dplyr::select(`geography code`, `2001 Owned`, `2001 Rented`, `2001 Social`)

# Ensure columns are numeric 2011
tenure_2011 <- tenure_2011 %>%
  mutate(
    `Owned outright` = as.numeric(`Owned outright`),
    `Owned with a mortgage or loan` = as.numeric(`Owned with a mortgage or loan`),
    `Shared ownership` = as.numeric(`Shared ownership`),
    `Rented from Local Authority` = as.numeric(`Rented from Local Authority`),
    `Other social rented` = as.numeric(`Other social rented`),
    `Private landlord or letting agency` = as.numeric(`Private landlord or letting agency`),
    `Other private rented` = as.numeric(`Other private rented`),
    `All Households` = as.numeric(`All Households`)
  )

# Perform the mutate operations
tenure_2011 <- tenure_2011 %>%
  mutate(
    `2011 Owned` = (`Owned outright` + `Owned with a mortgage or loan` + `Shared ownership`) / `All Households`,
    `2011 Social` = (`Rented from Local Authority` + `Other social rented`) / `All Households`,
    `2011 Rented` = (`Private landlord or letting agency` + `Other private rented`) / `All Households`
  )

# Keep only the required columns 2011
tenure_2011 <- tenure_2011 %>%
  dplyr::select(`LSOA code`, `2011 Owned`, `2011 Rented`, `2011 Social`)

# Ensure columns are numeric 2021
tenure_2021 <- tenure_2021 %>%
  mutate(
    `Owned outright` = as.numeric(`Owned outright`),
    `Owned with a mortgage or loan` = as.numeric(`Owned with a mortgage or loan`),
    `Shared ownership` = as.numeric(`Shared ownership`),
    `Rented from Local Authority` = as.numeric(`Rented from Local Authority`),
    `Other social rented` = as.numeric(`Other social rented`),
    `Private landlord or letting agency` = as.numeric(`Private landlord or letting agency`),
    `Other private rented` = as.numeric(`Other private rented`),
    `All Households` = as.numeric(`All Households`)
  )

# Perform the mutate operations
tenure_2021 <- tenure_2021 %>%
  mutate(
    `2021 Owned` = (`Owned outright` + `Owned with a mortgage or loan` + `Shared ownership`) / `All Households`,
    `2021 Social` = (`Rented from Local Authority` + `Other social rented`) / `All Households`,
    `2021 Rented` = (`Private landlord or letting agency` + `Other private rented`) / `All Households`
  )

# Keep only the required columns 2021
tenure_2021 <- tenure_2021 %>%
  dplyr::select(`LSOA code`, `2021 Owned`, `2021 Rented`, `2021 Social`)


# Join the datasets
tenure_2001 <- tenure_2001 %>%
  rename('LSOA code' = `geography code`)
tenure_2001 <- tenure_2001 %>%
  mutate(`LSOA code` = as.character(`LSOA code`))
tenure_2011 <- tenure_2011 %>%
  mutate(`LSOA code` = as.character(`LSOA code`))
tenure_2021 <- tenure_2021 %>%
  mutate(`LSOA code` = as.character(`LSOA code`))
joined_tenure <- tenure_2001 %>%
  full_join(tenure_2011, by = "LSOA code") %>%
  full_join(tenure_2021, by = "LSOA code")
cleaned_joined_tenure <- joined_tenure %>%
  drop_na()

# Interpolate missing years
years <- seq(2011, 2018)
# Process the data
interpolated_tenure <- cleaned_joined_tenure %>%
  pivot_longer(cols = -`LSOA code`, names_to = "variable", values_to = "value") %>%
  separate(variable, into = c("year", "category"), sep = " ") %>%
  mutate(year = as.numeric(year)) %>%
  complete(`LSOA code`, category, year = years) %>%
  group_by(`LSOA code`, category) %>%
  arrange(year) %>%
  mutate(value = zoo::na.approx(value, maxgap = Inf, na.rm = FALSE)) 

# Calculate at msoa level
lsoa_tenure <- lsoa %>%
  left_join(interpolated_tenure, by = c("LSOA11CD" = "LSOA code"))
# Select the required columns
selected_tenure <- lsoa_tenure %>%
  dplyr::select(MSOA11CD, category, year, value)

wide_tenure <- selected_tenure %>%
  pivot_wider(names_from = category,
values_from = value)

# get unique msoa values
library(dplyr)
# Grouping by MSOA and year, then calculating average tenure (Owned, Rented, Social)
grouped_data <- wide_tenure %>%
  group_by(MSOA11CD, year) %>%
  summarise(
    average_owned = mean(Owned, na.rm = TRUE),
    average_rented = mean(Rented, na.rm = TRUE),
    average_social = mean(Social, na.rm = TRUE)
  )

```

## Create a complete neighborhood dataset
```{r}
# Ensure the column names are consistent across all datasets
expanded_religion <- expanded_religion %>% rename(MSOA11CD = msoa11cd, Year = year)
grouped_data <- grouped_data %>% rename(Year = year)

# Ensure 'Year' is numeric (integer) across all datasets
expanded_religion <- expanded_religion %>% 
  mutate(Year = as.integer(Year))

interpolated_accommodation <- interpolated_accommodation %>% 
  mutate(Year = as.integer(Year))

grouped_data <- grouped_data %>% 
  mutate(Year = as.integer(Year))

msoa_crime <- msoa_crime %>% 
  mutate(Year = as.integer(Year))

msoa_green <- msoa_green %>% 
  mutate(Year = as.integer(Year))

# Join all datasets using full_join
neighborhood <- msoa_crime %>%
  full_join(msoa_green, by = c("MSOA11CD", "Year")) %>%
  full_join(interpolated_accommodation, by = c("MSOA11CD", "Year")) %>%
  full_join(expanded_religion, by = c("MSOA11CD", "Year")) %>%
  full_join(grouped_data, by = c("MSOA11CD", "Year"))

# Clean the resulting data by removing rows with any missing values
neighborhood_clean <- neighborhood %>%
  drop_na()

# Select relevant columns and remove duplicates
neighborhood_clean <- neighborhood_clean %>%
  dplyr::select(
    MSOA11CD,
    Year,
    MSOA_crime,
    MSOA_percent_green,
    geometry.x,        
    Occupation,
    religion,
    average_owned,
    average_rented,
    average_social
  ) %>%
  rename(geometry = geometry.x)  
```

```{r}
# Convert religion to a factor
neighborhood_clean <- neighborhood_clean %>%
  mutate(religion_factor = as.factor(religion))

# Convert the factor to numeric values
neighborhood_clean <- neighborhood_clean %>%
  mutate(religion_numeric = as.numeric(religion_factor))
```

## Incorporate other data
```{r}
# Preprocess housing data
housing_data_cleaned <- housing_data %>%
  mutate(year = as.integer(substr(date, 1, 4))) %>%
  dplyr::select(area, year, number_of_jobs, population_size, no_of_houses, median_salary) %>%
  na.omit()

# Verify numeric conversion for specific columns
housing_data_cleaned <- housing_data_cleaned %>%
  mutate(across(c(year, number_of_jobs, population_size, no_of_houses, median_salary), as.numeric))

# Filter housing_data_cleaned to keep only years 2011-2018
housing_data_filt <- housing_data_cleaned %>%
  filter(year >= 2011 & year <= 2018)

# Ensure all columns are numeric (except for the area name)
housing_data_filt <- housing_data_filt %>%
  mutate(
    number_of_jobs = as.numeric(number_of_jobs),
    population_size = as.numeric(population_size),
    no_of_houses = as.numeric(no_of_houses),
    median_salary = as.numeric(median_salary)
  )

# Standardize the 'area' and 'LAD11NM' for consistent matching
housing_data_filt <- housing_data_filt %>%
  mutate(area = tolower(area))
msoa <- msoa %>%
  mutate(LAD11NM = tolower(LAD11NM))

# Join to msoa dataset using LAD11CD (ensure LAD11CD is character in both datasets)
msoa_housing <- housing_data_filt %>%
  left_join(msoa, by = c("area" = "LAD11NM"))

# Calculate the proportion at MSOA level
housing_final <- msoa_housing %>%
  group_by(LAD11CD, year) %>%
  mutate(
    borough_population = sum(USUALRES, na.rm = TRUE),  # Total population per borough
    msoa_proportion = USUALRES / borough_population    # Proportion of MSOA population to the borough
  ) %>%
  ungroup()

housing_ff <- housing_final %>%
  mutate(
    number_of_jobs = number_of_jobs * msoa_proportion,
    population_size = population_size * msoa_proportion,
    no_of_houses = no_of_houses * msoa_proportion,
    median_salary = median_salary * msoa_proportion
  ) %>%
  dplyr::select(MSOA11CD, year, number_of_jobs, population_size, no_of_houses, median_salary)

# Select the final columns
housing_ff <- housing_ff %>%
  dplyr::select(MSOA11CD, year, number_of_jobs, population_size, no_of_houses, median_salary)
```

## Final join
```{r}
# Rename the 'year' column in housing_ff to 'Year'
housing_ff <- housing_ff %>%
  rename(Year = year)

# Perform the full join
neighborhood_house <- neighborhood_clean %>%
  full_join(housing_ff, by = c("MSOA11CD", "Year"))

# Clean the resulting data by removing rows with any missing values
neighborhood_house_clean <- neighborhood_house %>%
  drop_na()
```

## Incorporate borough and msoa information
```{r}
# Join the crime data with the shapefile data
combined_data <- neighborhood_house_clean %>%
  left_join(msoa_geometry %>% select(MSOA11CD, LAD11NM, LAD11CD), by = "MSOA11CD")

# Remove the geometry.y column
combined_data <- combined_data %>%
  dplyr::select(-geometry.y) %>%
  rename(geometry = geometry.x)
```

## Filter for specific 5 boroughs for analysis
```{r}
# View the unique borough names in the LAD11NM column
unique_boroughs <- unique(combined_data$LAD11NM)
# Print the unique borough names
print(unique_boroughs)

# Filter to select 5 boroughs (replace with actual borough names)
selected_boroughs <- c("Islington", "Hackney", "Tower Hamlets", "Camden", "Westminster")
bayesian_data <- combined_data %>% filter(LAD11NM %in% selected_boroughs)
```


## Correlation matrix
```{r}
# Correlation matrix

# Ensure the necessary library is loaded
library(dplyr)
library(corrplot)

# Remove non-numeric columns
numeric_data <- bayesian_data %>%
  dplyr::select(where(is.numeric))

# Calculate the correlation matrix using only numeric columns
correlation_matrix <- cor(numeric_data, use = "complete.obs")

# Print the correlation matrix
print(correlation_matrix)

# Visualize the correlation matrix using corrplot
corrplot(correlation_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         col = colorRampPalette(c("blue", "white", "red"))(200))
```

## EDA at 5 borough level
```{r}
summary_stats <- bayesian_data %>%
  group_by(Year) %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, median = median), na.rm = TRUE))
```

```{r}
library(ggplot2)
library(sf)
library(dplyr)

# Assuming your dataset is named bayesian_data and already contains spatial data
# Filter the data to include only the selected years
selected_years <- c(2011, 2014, 2016, 2018)
filtered_data <- bayesian_data %>%
  filter(Year %in% selected_years)

# List of variables to plot (excluding geometry, Year, MSOA11CD, LAD11NM, LAD11CD)
variables_to_map <- c("MSOA_crime", "MSOA_percent_green", "Occupation", "average_owned", 
                      "average_rented", "average_social", "number_of_jobs", "population_size", 
                      "no_of_houses", "median_salary")

# Loop through each variable and each year
for (var in variables_to_map) {
  for (year in selected_years) {
    # Filter the data for the specific year
    year_data <- filtered_data %>%
      filter(Year == year)
    
    # Generate the map
    p <- ggplot(year_data) +
      geom_sf(aes(geometry = geometry, fill = .data[[var]])) +
      scale_fill_viridis_c() +
      theme_minimal() +
      labs(title = paste("Map of", var, "in", year),
           fill = var) +
      theme(legend.position = "right")
    
    # Save the map as a PNG file
    ggsave(filename = paste0("map_", var, "_", year, ".png"), plot = p, width = 8, height = 6)
  }
}

```


## Prepare for bayesian model
```{r}
# Validate geography
bayesian_sf <- st_as_sf(bayesian_data, crs = 27700)

# Check for invalid geometries
invalid_geom <- st_is_valid(bayesian_sf, reason = TRUE)
print(invalid_geom)

# Check for invalid geometries and eliminate them
valid_bayesian_sf <- bayesian_sf %>%
  filter(st_is_valid(geometry))
```

## Make sure all msoas have 8 years
```{r}
# Define the range of years
required_years <- 2011:2018

# Filter out MSOAs that do not have data for all 8 years
bayesian_msoas <- bayesian_sf %>%
  group_by(MSOA11CD) %>%
  filter(all(required_years %in% Year)) %>%
  ungroup()
```

```{r}
# (if it's already loaded as sf, you can skip this part)
if (!inherits(bayesian_msoas, "sf")) {
  bayesian_msoas <- st_as_sf(bayesian_msoas, wkt = "geometry")
}

# Check unique years and MSOAs
unique_years <- unique(bayesian_msoas$Year)
unique_msoas <- unique(bayesian_msoas$MSOA11CD)
num_years <- length(unique_years)
num_msoas <- length(unique_msoas)
```

```{r}
# Select predictors
predictors <- c("religion_numeric", "MSOA_percent_green", "Occupation", "average_owned", 
                "average_rented", "average_social", "number_of_jobs", "population_size", 
                "no_of_houses", "median_salary")

# Convert predictors to a numeric matrix, ensuring geometry is dropped
X_matrix <- bayesian_msoas %>%
  st_drop_geometry() %>%  # Drop the geometry column
  dplyr::select(religion_numeric, Occupation, average_owned, 
                average_rented, population_size, 
                no_of_houses, median_salary) %>%  # Explicitly select the predictors
  mutate(across(everything(), as.numeric)) %>%  # Ensure all columns are numeric
  as.matrix()  # Convert to a matrix

# Verify the structure of X_matrix to ensure it is now a numeric matrix without geometry
str(X_matrix)

# Prepare the response variable
y <- as.numeric(bayesian_msoas$MSOA_crime)
```

```{r}
# Calculate neighbors based on centroids and a suitable distance
coords <- st_centroid(st_geometry(single_year_data))
neighbors_distance <- dnearneigh(coords, 0, 1500)  # Set a distance threshold

# Create spatial weights matrix
W_distance <- nb2mat(neighbors_distance, style = "B", zero.policy = TRUE)

# Visualize the connections to verify
plot(st_geometry(single_year_data), border="blue")
plot(neighbors_distance, coords, add=TRUE, col="red")
```
```{r}
# Convert to spatial weights matrix
W_distance <- nb2mat(neighbors_distance, style = "B", zero.policy = TRUE)

# Print the matrix to check non-zero values
print(W_distance)
```

## Stan model
```{r}
# Check if there are any non-integer values in y
any_non_integers <- any(bayesian_msoas$MSOA_crime != floor(bayesian_msoas$MSOA_crime))
print(any_non_integers)

# If there are non-integer values, convert y to integers
y <- as.integer(round(bayesian_msoas$MSOA_crime))

```

```{r}
X_matrix_scaled <- scale(X_matrix)

# Prepare the data for Stan
stan_data <- list(
  N = nrow(bayesian_msoas),          # Number of observations
  K = ncol(X_matrix),                # Number of predictors (10)
  y = y,                             # Response variable
  X = X_matrix_scaled,                      # Predictor matrix
  W = as.matrix(W_distance),         # Updated to use the distance-based weight matrix
  num_msoas = num_msoas,             # Number of MSOAs
  num_years = num_years              # Number of years
)

# Check structure of stan_data to ensure correctness
str(stan_data)

# Fit the model using Stan
fit <- stan(
  file = "C:/Users/USUARIO/Documents/UCL/Summer/bayesian/model.stan",
  data = stan_data,
  chains = 4,
  iter = 6000,  # Increase iterations
  control = list(max_treedepth = 20, adapt_delta = 0.99) 
)

# Model diagnostics and output
print(fit, pars = c("beta", "alpha", "sigma_phi"))
traceplot(fit, pars = c("beta", "alpha", "sigma_phi"))
summary(fit)
pairs(fit)

# Cross-validation
loo_results <- loo(fit)
print(loo_results)
```

## Map of predicted crime
```{r}
# Extract the predicted crime rates from the Stan model
predicted_crimes <- extract(fit)$y_pred
bayesian_msoas$predicted_crime <- apply(predicted_crimes, 2, mean)

# Create a map of predicted crime rates
ggplot(data = bayesian_msoas) +
  geom_sf(aes(fill = predicted_crime)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Predicted Crime Rates by MSOA",
       fill = "Predicted Crime Count")
```

## Map of exceedence probability
```{r}
# Set the threshold for exceedance
threshold <- 30

# Calculate the exceedance probability for each MSOA
bayesian_msoas$exceedance_prob <- apply(predicted_crimes > threshold, 2, mean)

# Create a map of exceedance probabilities
ggplot(data = bayesian_msoas) +
  geom_sf(aes(fill = exceedance_prob)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Exceedance Probability of Crime Rates",
       fill = "Probability")
```

## Comparison map
```{r}
# Calculate the difference between actual and predicted crime rates
bayesian_msoas$crime_diff <- bayesian_msoas$MSOA_crime - bayesian_msoas$predicted_crime

# Create a comparison map showing actual vs predicted crime rates
ggplot(data = bayesian_msoas) +
  geom_sf(aes(fill = crime_diff)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(title = "Difference Between Actual and Predicted Crime Rates",
       fill = "Actual - Predicted")
```



